---
title: "Conversion-Risk Diagnostic: Gross vs Net — Handoff"
subtitle: "PPR Acquisition Policy Under Nonstationarity"
author: "Session summary for a future AI agent"
date: "2026-08-03"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# How to use this document

This is a **handoff record**, not a runnable analysis. All chunks are
`eval = FALSE`. It summarises one working session (2026-08-03) whose purpose was
to diagnose whether the conversion-risk variable in this project measures what it
is supposed to measure.

**The short answer: it does not, and the diagnostic quantifies by how much.**

Read §2 for the finding, §5 before touching the raster code, and §7 for what is
still open.

---

# 1. Project context

Graduate thesis project asking whether accounting for non-stationarity improves
land-acquisition decisions for waterfowl conservation in the U.S. Prairie Pothole
Region.

Three policies acquire parcels decade by decade (2020–2100) under a budget
constraint (5 median-priced EAUs per period, no rollover):

| Policy | Foresight | Notes |
|---|---|---|
| Greedy | None | Rank by current benefit/cost, buy down the list |
| Myopic ILP | Fabricated stationary future | Freezes current benefit + hazard, re-solves each decade |
| Rolling-horizon ILP | True projected trajectories | Near-optimal benchmark; Gurobi, MIP gap ~0.01% |

Acquisition value is expected loss prevented:

$$V_{i,\tau} = \sum_{t=\tau}^{T} \delta^t \, b_{i,t} \, (1 - S_{i,t})$$

**Dimensions:** 879 EAUs across 20 WMDs (after dropping WMDs missing >20% of EAU
land-cover data), 9 decadal periods, 5 scenarios (RCP 4.5/8.5 × wet/dry + a
stationary null). Discount factor 0.95 per decade.

**Grid facts established this session** (worth knowing before any raster work):

- EAU grid: 46 × 85 = 3,910 cells at **18,514 m** resolution
- `scf` = 380,834 native FOREsce pixels per EAU cell
- 1,210 EAUs in the lookup; **1,169** have land-cover coverage; **879** survive
  into the panel after WMD drops
- Native raster ≈ 1.49e9 pixels ≈ 1.4 GB per INT1U layer

## 1.1 Pipeline

| Script | Role |
|---|---|
| `00_setup.R` | Paths, `HAS_SPATIAL`, `DIR_DERIVED`, `DIR_FIGS` |
| `01_create_EAUs.R` | EAU grid, `eau_wmd_lookup.csv` |
| `02_EAU_prop_suitable.R` | `prop_suitable` per EAU-year from FOREsce; `foresce_eau_shared_mask.tif` |
| `03_create_data_panel.R` | Panel assembly, by-cell-number join, wet/dry duplication, stationary rows |
| `04_benefit_data.R` | McKenna WMD abundance distributed to EAUs by habitat share |
| `05_risk_of_loss.R` | **`trans_prob`** — the variable under investigation |
| `06_cost_data.R` | Nolte land values + 2%/yr |
| `07_ilp_core.R` | ILP engine, `evaluate_policy()` |
| `08_run_models.R` | Model runs |
| `09_validation.R` | Validation |
| `10`–`11` | Results assets, Quarto report |
| `12_diagnostic_gross_vs_net.R` | **This session's work** |

## 1.2 The structural coupling to remember

`prop_suitable` is load-bearing **twice**: it sets the hazard (script 05) *and*
the within-WMD benefit allocation (`scaled_abundance = habitat_share ×
abs_abundance`, script 04). Any change to how the FOREsce stock series is read
propagates into both arms of $V$. This is why the 7/30 spatial-join fix moved
results as much as it did, and it constrains any redesign of the hazard.

---

# 2. THE FINDING

## 2.1 The problem

Script 05 computes the hazard as a **net change in habitat stock**:

```{r}
trans_raw <- (ps_t - ps_t1) / ps_t
trans_prob <- pmax(epsilon, trans_raw)
```

Writing $L$ for area converting suitable → unsuitable and $G$ for the reverse:

$$ps_{t+1} = ps_t - L + G \;\Longrightarrow\; \frac{ps_t - ps_{t+1}}{ps_t} = \frac{L - G}{ps_t}$$

So script 05 computes $(L-G)/ps_t$ where $L/ps_t$ was intended. The two agree only
where $G = 0$. Recovering $L$ and $G$ separately requires returning to pixel
level, because the aggregation in script 02 nets them irreversibly.

## 2.2 Headline results (full run, 16 year-pairs, 18,704 EAU-years)

**The net metric conceals 22.8% of all habitat conversion.**

| Landscape total (EAU-fraction units, both RCPs, 2020–2090) | Value |
|---|---|
| Gross loss $L$ | 40.72 |
| Gross gain $G$ | 9.27 |
| Net ($L-G$) | 31.45 |
| **Hidden ($G/L$)** | **22.8%** |

Units note: these are **areas, not percentages**. `prop_lost` is the fraction of
an EAU's *total* area (denominator `scf`) converting in that decade; summing over
EAU-years gives "EAU-equivalents of area." The sum spans both RCPs, so halve for
a single scenario.

| Censoring (non-terminal EAU-years) | Count | Share |
|---|---|---|
| Total rows | 18,688 | — |
| With any real loss (gross > 0) | 14,477 | 77.5% |
| **Censored (gross > 0, net ≤ 0)** | **1,659** | **8.9% of all, 11.5% of losing** |
| Understated (gross > net) | 11,110 | 59.4% |
| Gross loss exactly zero | 4,211 | 22.5% |

Censored-row magnitudes: mean 0.00227, median 0.00069, p90 0.00619, max 0.03555.
The worst censored parcel loses **3.6% of its habitat in a decade** and enters the
model as risk-free.

## 2.3 The 2050 dip is REAL, not a netting artefact

This closes the "(currently undiagnosed) drop in 2050" note in the .qmd.

Gross loss collapses at 2050 as well as net — mean per-decade conversion falls
from **1.87% (2040) to 0.20% (2050)**, a ~9× drop, in both RCPs. Since gross loss
cannot be masked by restoration, the drop is in the FOREsce data itself: the GCAM
reference scenario front-loads essentially all cropland expansion before 2050.
Consecutive-year `prop_suitable` correlations run 0.99956–0.99997 with none
flagged, ruling out duplicate raster layers.

## 2.4 But netting does serious damage mid-century

| Decade | gross | net | ratio | % censored |
|---|---|---|---|---|
| 2020–2040 | ~0.018 | ~0.017 | 1.02–1.27 | 0.5–7% |
| **2050–2070 (4.5)** | ~0.002 | 0.0009 → **−0.0003** | 2.2 → 4.5 → **NA** | **14–17%** |
| **2050–2070 (8.5)** | ~0.002 | **−0.0033, −0.0009, −0.0013** | **all NA** | **16–24%** |
| 2080–2090 | ~0.003 | ~0.002 | ~1.27 | ~4% |

In **four decade-scenario cells the mean net loss is negative** — the landscape
reads as net-restoring, so essentially the whole decade floors to epsilon despite
~0.2%/decade of real conversion.

## 2.5 A scenario-ordering inversion

At 2050, RCP 8.5 has ~5× the restoration of 4.5 (`gain_mean` 0.00538 vs 0.00110):

| | gross loss | gain | net |
|---|---|---|---|
| RCP 4.5 | 0.00204 | 0.00110 | **+0.00094** |
| RCP 8.5 | 0.00206 | 0.00538 | **−0.00332** |

Habitat is destroyed at effectively the **same rate** in both (0.204% vs 0.206%).
The entire net gap comes from restoration. The model is therefore told that
high-emissions RCP 8.5 carries *less* conversion risk than RCP 4.5.

**Why this is an artefact, not a finding:** acquisition prevents conversion — that
is the causal mechanism $V$ rests on. Restoration elsewhere in an EAU does not
reduce the risk to habitat currently there, and buying a parcel does not cause the
gain. Netting makes the hazard answer "is this EAU's stock shrinking?" when the
decision needs "is this EAU's existing habitat at risk?"

This likely already shows in Table 1, where RCP 8.5 has lower value-over-baseline
than 4.5 in every policy. Some of that gap is probably mid-century hazard being
floored away by restoration that acquisition neither prevents nor captures.

## 2.6 Epsilon

The user's prior was that the data-derived `epsilon` in script 05 exists only
because the net metric admits growth, and becomes obsolete under a flow metric.
**Confirmed, and more strongly than argued:** `epsilon <- min_positive_loss / 10`
is *constructed* to sit below every positive loss value, so
`{trans_raw < epsilon}` and `{trans_raw <= 0}` select the **same rows**. Testing
censoring against 0 is exactly equivalent. (The only divergence would be
`epsilon_override`, which is `NULL`.)

Under a flow metric `gross_loss >= 0` identically, so nothing can go negative.
**But** 4,211 rows (22.5%) have gross loss of exactly zero — genuinely no
conversion, not netted away. A background-risk decision is still required, just an
honest one (a parameter chosen and defended) rather than a patch for an impossible
value.

---

# 3. Deliverables produced

| File | Status | Purpose |
|---|---|---|
| `12_diagnostic_gross_vs_net.R` | Modified, **not re-run since final fix** | Main diagnostic |
| `12b_recover_finish.R` | Run successfully | Finished sections B–D from a live session after a crash |
| `12c_patch_sentinel.R` | **Written, not confirmed run** | Repairs 16 sentinel-collision rows |

All diagnostic output goes to `diag_gross_v_net/` only — deliberately *not*
`derived_data/` or `output_figs/`, so it can never be mistaken for canonical
pipeline data:

- `diag_gross_vs_net.csv` (18,688 rows; 18,704 after patch)
- `diag_summary.csv` (16 rows: rcp × year)
- `diag_gross_vs_net_scatter.png`
- `progress.log`

`RUN_TAG` appends `_QUICK` on smoke tests so partial artefacts are
self-identifying and cannot overwrite a real run.

---

# 4. Changes made to script 12 (and why)

## 4.1 User-requested

1. **Isolated output folder** `diag_gross_v_net/`.
2. **Scatter figure only** — time-series figure dropped.
3. **Literature-comparison section removed** (was section D); anomaly probe
   renumbered E → D. `LIT_RATES` deleted.
4. **Censoring tested against 0**, not epsilon (see §2.6 — equivalent by
   construction).
5. **`ratio` returns `NA`** when `net_mean <= 0` instead of
   `pmax(net_mean, 1e-12)`, which silently converted negative denominators to
   1e-12 and printed ~1e7. Added `gross_minus_net`, always well-defined.
6. **Class decomposition deferred** by user decision (see §7).

## 4.2 Performance work

Runtime went from a projected **5.5 h serial** to **23.8 min**:

| Change | Effect |
|---|---|
| Serial one-slot raster cache | 24% — **later removed**, incompatible with parallel dispatch |
| Parallel over 16 year-pairs (PSOCK) | ~4.4× |
| Single-pass `lapp` (12 native passes → 3) | ~4.1× per pair |

The single-pass rewrite is the important one. Per pair it now does:

1. **One** `crop` over a 2-layer stack of both source years
2. **One** `lapp` emitting all flow layers at once — complements and products are
   transient vectors inside the chunk callback, never separate rasters
3. **One** multi-layer `resample` (replacing four separate calls)

Temp I/O: ~56 GB → ~12 GB per pair.

**Deliberately NOT done** (do not "optimize" these later):

- `resample(method = "sum")` is **not** replaced by `aggregate(fact = ...)`.
  It would be far faster but would no longer reproduce script 02's aggregation
  exactly, and check A2 is the entire basis for comparing to the pipeline.
- `ps_t1` is **not** derived as `ps_t - L + G`. That would save a resample but
  make check A1 true by construction.

## 4.3 Robustness added

- **Scratch-space preflight** — computes requirement from `n_cells × scf`, reads
  free space via `df -Pk`, stops *before* a long run rather than during one.
  `TERRA_TEMPDIR` redirects terra's scratch.
- **`TERRA_MEMFRAC`** — auto-set to `0.6 / n_workers` and applied *inside* the
  worker. terra's `memfrac` is per-process and workers inherit it; left at
  default, N workers can commit N × 60% of RAM and swap. On macOS that presents
  as slowness, not an error.
- **`PARALLEL_BACKEND`** — `"psock"` (default, RStudio-safe), `"fork"`
  (Rscript only), `"serial"`.
- **Progress log** — PSOCK workers cannot print to the RStudio console, so
  per-pair completions append to `diag_gross_v_net/progress.log`.
- **`compareGeom` guard** — fails loudly with both filenames if source years
  don't share a grid, rather than mis-registering loss against gain.

---

# 5. BUGS FOUND — read before touching the raster code

Six bugs surfaced. Four were introduced during this session. Recorded honestly
because the failure *modes* generalise.

## 5.1 INT1U sentinel collision (most important)

**Symptom:** 16 EAU-years (15 `prop_lost`, 1 `prop_gained`) came back `NaN` while
`ps_t` and `ps_t1` were clean. Not edge cells — `ps_t` ran 0.08–0.70, median 0.44
— and perfectly symmetric across RCPs.

**Cause:** the flow raster was written as `INT1U`, whose nodata sentinel is **255**.
The resampled output **inherits that nodata value**, so any EAU whose summed flow
count landed on exactly 255 pixels was read back as `NaN` rather than 255.

**Confirmation:** every affected row reconstructs via the stock identity to
exactly `255/scf` = 0.0006695831. `ps_t`/`ps_t1` were untouched only because their
sums (3e4–3e5) can never equal 255.

**Fix:** no NA is ever written to the raster. All four flow layers are strict 0/1;
coverage travels as a **fifth 0/1 layer** whose sum gives covered pixel count per
EAU. Missingness is applied at EAU resolution *after* aggregation, where no
sentinel exists. `NAflag = -1` on the resample as belt-and-braces (sums are
non-negative, so no collision possible).

```{r}
# CURRENT (fixed) form — note: no NA written, coverage as a count layer
flow <- terra::lapp(s, fun = function(a, b) {
  ok  <- !is.na(a) & !is.na(b)
  st  <- as.integer(a %in% classes & ok)
  st1 <- as.integer(b %in% classes & ok)
  cbind(st, st1,
        st * (1L - st1) * as.integer(ok),   # lost
        (1L - st) * st1 * as.integer(ok),   # gained
        as.integer(ok))                     # coverage
}, wopt = list(datatype = "INT1U"))
```

> **GENERAL RULE:** never encode missingness in a value that a legitimate sum
> could produce. This applies to any integer-typed intermediate elsewhere in the
> pipeline, including script 05 if the hazard is reworked.

## 5.2 `na.rm = TRUE` masked the check meant to catch 5.1

Check A1 used `max(abs(ident_resid), na.rm = TRUE)`, so it **skipped exactly the
rows that were broken** and reported PASS. Now counts NA residuals and FAILS if
any exist.

## 5.3 Filtering on `ps_t` alone

`filter(!is.na(ps_t))` let through rows where `ps_t` was present but a partner
column was NA. These propagated into `gross_loss`, turned every `sum()` into NA,
and crashed section B at `if (cens$n_censored > 0)`. Now filters on all flow
columns and reports the two failure modes separately (expected no-coverage vs
unexpected partial decomposition).

## 5.4 `%in%` returns FALSE for NA, not NA

In the `lapp` callback, a no-coverage pixel would have silently become 0 and
entered the sums as real unsuitable land. Caught before it shipped.

## 5.5 `on.exit()` at top level

Written for PSOCK cluster teardown, but `on.exit` registers against a *function*
frame and there is none in a sourced script — it would have fired immediately and
stopped the cluster before any work ran. Replaced with
`tryCatch(..., finally = stopCluster(cl))`.

## 5.6 Backend failure semantics differ

`mclapply` returns worker failures as list elements; `parLapplyLB` **raises**.
Under PSOCK the per-pair failure report would never have fired, and one failure at
hour two would have discarded all completed pairs. The worker call is now wrapped
in `try()` so both backends behave identically.

## 5.7 Figure title was inverted (pre-existing)

Said "Every point **below** the dashed line understates conversion risk." Since
`gross − net = gain/ps_t ≥ 0` identically, **no point can fall below the
diagonal**. Corrected to "above," with a caption noting the line doubles as a
visual check on the decomposition.

---

# 6. Run environment and performance

macOS, 12 cores detected, ~790 GB free on the boot volume. terra scratch defaults
to `/var/folders/...` on `/`.

**Settings used for the successful full run:** `QUICK <- FALSE`,
`N_CORES <- 6L`, `PARALLEL_BACKEND <- "psock"`, `TERRA_TEMPDIR <- NULL`,
`TERRA_MEMFRAC <- NULL` (auto → 0.100).

**Result:** 23.8 min wall, 105.3 min summed worker time, **4.4× speedup**.

## 6.1 I/O contention is measurable

Same rasters, same code — only worker count varies:

| Concurrent workers | Per-pair time |
|---|---|
| 2 | ~311 s |
| 6 | ~405–461 s |
| 3–4 (final group) | ~296–314 s |

Six workers cost ~45% more per pair than two. Throughput still won, but this is
why the speedup was 4.4× not 6×. **The job is disk-bound, not CPU-bound.**

## 6.2 Scheduling

16 pairs ÷ 6 workers leaves a 5-minute tail where one pair ran with five workers
idle (~17% of capacity). **Use `N_CORES <- 8L` (2 clean waves) or `4L` (4 waves)**
next time. Utilisation was 73.7%.

Per-pair costs are uniform (meanclim-anchored pairs cost the same as the rest), so
`parLapplyLB`'s load balancing buys little here; prescheduled dispatch would
perform identically.

## 6.3 Estimation lesson

Runtime was mis-estimated **three times**, twice optimistically. The root error:
script 02's "~4 min per scenario" comment is **parallel wall-clock across all
cores**, not serial cost per raster. Script 02 also does far less per unit (1
classify + 1 resample per file, no native-resolution intermediates).

> Calibrate from measured per-pair timings in `progress.log`, not from comments in
> other scripts.

## 6.4 Production cost, if the metric is adopted

A production gross-loss pipeline would be **~2.3× script 02**, not 15×. The
diagnostic is expensive because it computes $G$ (production needs only $L$),
resamples `ps_t1` independently (to keep A1 honest), and cannot reuse rasters
across pairs (parallelism replaced the cache). Fold `prop_lost` into script 02's
existing loop rather than adding a script. It is not in the iteration loop — 05 →
08 stay fast.

---

# 7. OPEN ITEMS

## 7.1 Immediate

1. **Confirm `12c_patch_sentinel.R` ran.** Written but not verified. It
   re-verifies each row reconstructs to exactly 255 px (and *stops* if not), fills
   `255/scf`, recomputes derived columns, re-runs A1 with no `na.rm` on all rows,
   and re-saves with a `patched_sentinel` flag. Effect on conclusions is nil
   (~0.011 against L = 40.7).
2. **Script 12 has not been re-run since the §5.1 fix.** Current CSVs come from
   the pre-fix run plus recovery. Results are correct after patching, but a clean
   re-run would be tidier (~16 min at `N_CORES <- 8L`).
3. **Stale comment:** the RUNTIME header says "45–75 min at N_CORES = 4," written
   before measurement. Actual is ~20 min at 4 workers.

## 7.2 The main decision

**Switch `trans_prob` in script 05 to `prop_lost / ps_t` and re-run the models.**

Predictions to test:

- Absolute value-added rises (mid-century hazard stops being ~zero)
- Rolling-vs-myopic gap likely **widens in later periods**, where the two metrics
  diverge most
- Cross-scenario differences **compress** — some of RCP 8.5's lower
  value-over-baseline in Table 1 is metric artefact (§2.5)

Requires a background-risk decision for the 4,211 zero-loss rows (§2.6).

## 7.3 Deferred by user (not rejected)

- **Class decomposition.** `L` is currently one number, but literature rates are
  class-specific (wetland 0.57%/yr, grassland 1.33%/yr) and the suitable set lumps
  grass (19, 26) with water/wetland (20, 28, 29). No like-for-like comparison
  without splitting.
- **Destination decomposition.** Loss to cropland is conversion; open-water →
  barren under a dry FOREsce draw is hydrology. Currently identical numbers.
  Bears on whether the 2050 collapse is agricultural or hydrologic.

## 7.4 Discrepancies to reconcile

- **EAU area.** Grid is 18,514 m → **342.8 km²** (34,275 ha) per EAU. The project
  doc and Figure 1 caption say **~282 km²** (28,200 ha) — a 21% difference that
  propagates into any hectare figure. Cost-per-hectare is unaffected (equal-area
  cells, constant rescale); total-area claims are not.
- **Wet/dry land cover.** FOREsce has no wet/dry distinction, so land cover (and
  therefore conversion risk and within-WMD duck allocation) is constant across
  wet/dry. Only abundance varies. Pre-existing, documented limitation; interacts
  with the counter-intuitive wet-vs-dry gap in the .qmd.

## 7.5 Reporting suggestion

Quote the **area-weighted** landscape rate, not the mean of per-parcel ratios:

```{r}
nonterm %>% summarise(
  habitat_eau_years = sum(ps_t),
  loss_rate = sum(prop_lost)   / sum(ps_t),
  gain_rate = sum(prop_gained) / sum(ps_t))
```

(`gross_mean` = 0.84%/decade is the *unweighted* per-EAU mean.)

---

# 8. Notes for the next agent

**On this codebase**

- Script 12 is standalone and read-only w.r.t. the pipeline. Safe to run; safe to
  delete `diag_gross_v_net/`.
- Fidelity checks A1–A3 are the safety net. **A2** (recomputed `prop_suitable`
  matches panel) is the strongest — it is what licenses comparing diagnostic
  numbers to pipeline numbers. Full-run values: A1 7.56e-08; A2 5.55e-16
  (n=12,306); A3 4.53e-14 (n=8,045).
- A1's residual is ~8 orders larger than A2's for a benign reason: A1 sums four
  separately-rounded resampled layers so float32 rounding shows, while A2 compares
  the same quantity computed the same way twice, so rounding cancels.
- `QUICK <- TRUE` gives 2 pairs, RCP 4.5 only. **Sections B–D are not
  interpretable under QUICK** — it retains the two highest-conversion decades and
  cannot see 2050. It reported 4.2% concealment vs 22.8% on the full run.
- PSOCK workers receive only serialisable arguments (paths, vectors, numbers).
  **A `SpatRaster` cannot cross a PSOCK connection** — the EAU mask is re-opened
  from its path inside the worker. Preserve this if refactoring.

**On working with this user**

- Technically strong; catches things. The sentinel bug surfaced because they
  pushed on 16 anomalous rows rather than accepting "0.086%, probably edge
  effects."
- Asks direct clarifying questions when an explanation is unclear. Answer the
  actual confusion rather than restating.
- Explicitly invited push-back ("Push back if you disagree") — and was right on
  the epsilon question when I was wrong.
- Wants deferred items *noted*, not silently dropped.
- Do not claim verification that did not happen. R was unavailable in the agent
  environment all session; every claim about behaviour was static reasoning until
  the user ran it. Say so.
